---
title: "Math Handout"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library(lattice)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(grid)
library(gridExtra)
library(latex2exp)
library(tidyr)
library(broom)

options(lifecycle_disable_warnings = TRUE)

prDes <- readRDS("data/GSE4051_design.rds")
levels(prDes$devStage)[levels(prDes$devStage)=="4_weeks"] <- "4W"

prDat<-read.table("data/GSE4051_data.tsv",
                      sep = "\t", header = T, row.names = 1)

## I've selected this as our hit
theHit <- which(rownames(prDat) == "1440645_at") # 17843
## and this as our boring gene
theBore <- which(rownames(prDat) == "1443184_at") # 18898

keepers <- data.frame(row = c(theBore, theHit),
                       probesetID = I(rownames(prDat)[c(theBore, theHit)]))

devDat <- as.vector(t(prDat[keepers$probesetID, ]))
devDat <- data.frame(gene = rep(c("theBore", "theHit"), each = nrow(prDes)),gExp = devDat)
devDat <- data.frame(prDes, devDat)

boreDat <- filter(devDat, gene == "theBore")
hitDat <- filter(devDat, gene == "theHit")
```

## Two-way ANOVA or a linear model with interaction 

Which group means are we comparing in a model with 2 factors? 

For simplicity, we first consider only two levels of `devStage`: E16 and 4W

```{r,echo=FALSE, include=FALSE}
##########################################################
## simplying devStage to first and last timepoints
##########################################################
prDes <- 
  droplevels(subset(prDes,
                    subset = devStage %in%
                      levels(devStage)[c(1, nlevels(devStage))]))
str(prDes) # 15 obs. of  4 variables
prDat <- subset(prDat, select = prDes$sidChar)

# Gene selected for illustration
(luckyGene <- which(rownames(prDat) == "1455695_at")) # 26861
twoDat <- data.frame(gExp = unlist(prDat[luckyGene, ]))
twoDat <- data.frame(prDes, twoDat)
twoDat$grp <- with(twoDat, interaction(gType, devStage))
str(twoDat)
with(twoDat, table(gType, devStage))
table(twoDat$grp)
```


```{r,echo=FALSE,fig.height= 4, dev='svg'}
mu.hat<-twoDat %>% group_by(grp) %>% summarize(mean(gExp)) %>% as.data.frame()
lucky <- ggplot(twoDat, aes(x = devStage, y = gExp, group=gType, colour=gType)) + 
             geom_jitter(width = 0.1) +
             theme_bw() +
             theme(legend.justification=c(1,1),legend.position = c(1,1)) +
             ylim(6, 13) + stat_summary(aes(group=gType,colour=gType), fun.y=mean, geom="line",size=1.5) 

lucky +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
  geom_text(aes(x = .8, y = mu.hat[1,2], label = TeX("$\\widehat{\\mu_1}$", output = "character")),colour="black", size=6,parse = TRUE)+
  geom_text(aes(x = .8, y = mu.hat[2,2], label = TeX("$\\widehat{\\mu_2}$", output = "character")), colour="black", size=6,parse = TRUE)+
  geom_text(aes(x = 2.2, y = mu.hat[3,2], label = TeX("$\\widehat{\\mu_3}$", output = "character")), colour="black", size=6, parse = TRUE)+
  geom_text(aes(x = 2.2, y = mu.hat[4,2], label = TeX("$\\widehat{\\mu_4}$", output = "character")), colour="black", size=6, parse = TRUE)
```

$$\mu_1=E[Y_{(wt,E16)}], \ \mu_2=E[Y_{(NrlKO,E16)}], 
\ \mu_3=E[Y_{(wt,4W)}], \ \mu_4=E[Y_{(NrlKO,4W)}]$$

### Reference-treatment effect parametrization

By default, `lm` assumes a <font color = "red">**reference-treatment effect**</font> parametrization. We just need *more* dummy variables!!

### Mathematically (a bit more difficult...)

$$Y_{ijk}=\theta + \tau_{KO} \times x_{KO,ijk}+ \tau_{4W} \times x_{4W,ijk}+ \tau_{KO4W} \times x_{KO,ijk}\times x_{4W,ijk}+ \varepsilon_{ijk}$$
Subscripts: $i$ indexes samples per group, $j=\{wt, NrlKO\}$, $k=\{E16, 4W\}$  

The names of these parameters and variables look overwhelming but think of them as just names for:

> $x_{KO,ijk}$: a dummy variable with value 1 for NrlKO genotype samples (any sample with j=NrlKO), and 0 otherwise. I call this variable $x_{KO}$

> $x_{4W,ijk}$: a different dummy variable with value 1 for 4W  samples (any sample with k=4W), and 0 otherwise. I call this variable $x_{4W}$

> $\tau_{KO}$, $\tau_{4W}$, and $\tau_{KO4W}$: parameters to model the *simple* effects of genotype (NrlKO), development (4W), and their interaction

**Note**: in this "simple" version with 2 levels per factor we need only one dummy variable per factor: $x_{KO}$ and $x_{4W}$. But this model can be extended to multiple factors with multiple levels!! you just need more dummy variables!

### Reference: wt & E16

As before, comparisons are relative to a reference but now we have reference levels in both factors: E16 and wt

$$Y_{ijk}=\theta + \tau_{KO} \times x_{KO,ijk}+ \tau_{4W} \times x_{4W,ijk}+ \tau_{KO4W} \times x_{KO,ijk}\times x_{4W,ijk}+ \varepsilon_{ijk}$$
  
  
For any sample $i$ in the reference group: $j=wt$ and $k=E16$, then $x_{KO}=0$ and $x_{4W}=0$ (I'm omiting subscripts for clarity). Then only $\theta$ remains and we get:  $$E[Y_{wt,E16}]=\theta$$

as before $\theta$ is the mean of the reference group 


```{r,echo=FALSE,fig.height= 2.7, dev='svg'}
lucky +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
  geom_text(aes(x = .8, y = mu.hat[1,2], label = TeX("$\\widehat{\\theta}=\\widehat{\\mu_1}$", output = "character")),colour="black", size=6, parse = TRUE) + 
  geom_point(aes(x = 1, y = mu.hat[1,2]),size=7,shape=1,colour='#00BFC4')
```


```{r,tidy.opts=list(width.cutoff=40)}
twoFactFit <- lm(gExp ~ gType * devStage, twoDat)
summary(twoFactFit)$coeff
```

And the **`lm` estimate**, $\hat{\theta}$, is the sample mean of the group 

```{r,tidy.opts=list(width.cutoff=40)}
means.2Fact <- as.data.frame(twoDat %>% group_by(grp) %>% summarize(cellMeans=mean(gExp)))
(means.2Fact <-means.2Fact %>%
          mutate(txEffects=cellMeans-cellMeans[1], lmEst=summary(twoFactFit)$coeff[,1]))
```


### *Simple* genotype effect: wt *vs* NrlKO <font color="blue"> at E16 </font>


$$Y_{ijk}=\theta + \tau_{KO} \times x_{KO,ijk}+ \tau_{4W} \times x_{4W,ijk}+ \tau_{KO4W} \times x_{KO,ijk}\times x_{4W,ijk}+ \varepsilon_{ijk}$$
  
For any wt sample at E16: $x_{KO}=0$ and $x_{4W}=0$. Then <font color="red"> $E[Y_{wt,E16}]=\theta$ </font>
  
For any KO sample at E16: $x_{KO}=1$ and $x_{4W}=0$. Then <font color="red"> $E[Y_{NrlKO,E16}]=\theta + \tau_{KO}$ </font>
  
Substracting these expectations we get $\tau_{KO}$, the *conditional* genotype effect <font color="blue"> at E16 </font>: 
  
$$\tau_{KO}=E[Y_{NrlKO,E16}]-E[Y_{wt,E16}]$$ 
And its **`lm` estimate**, $\hat{\tau}_{KO}$, is the *difference* of sample respective means (check below)

```{r,echo=FALSE,fig.height= 2.7, dev='svg'}
lucky +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
  geom_text(aes(x = .7, y = abs(mu.hat[1,2]+mu.hat[2,2])/2, label = TeX("$\\widehat{\\tau_{KO}}$", output = "character")),colour="black", size=6, parse = TRUE) + 
  geom_point(aes(x = 1, y = mu.hat[1,2]),size=7,shape=1,colour='#00BFC4')+
  geom_point(aes(x = 1, y = mu.hat[2,2]),size=7,shape=1,colour='#F8766D')+
  geom_segment(aes(x=.8,y=mu.hat[1,2],xend=.8,yend=mu.hat[2,2]),colour=1,arrow = arrow(length = unit(0.1,"cm")))
```

```{r}
summary(twoFactFit)$coeff
means.2Fact
```
  
### *Simple* developmental effect: E16 *vs* 4W <font color="blue"> at wt </font>

$$Y_{ijk}=\theta + \tau_{KO} \times x_{KO,ijk}+ \tau_{4W} \times x_{4W,ijk}+ \tau_{KO4W} \times x_{KO,ijk}\times x_{4W,ijk}+ \varepsilon_{ijk}$$
  
For any wt sample at E16: $x_{KO}=0$ and $x_{4W}=0$. Then <font color="red"> $E[Y_{wt,E16}]=\theta$ </font>
  
For any wt sample at 4W: $x_{KO}=0$ and $x_{4W}=1$. Then <font color="red"> $E[Y_{wt,4W}]=\theta + \tau_{4W}$ </font>
  
Substracting these expectations we get $\tau_{4W}$, the *conditional* development effect <font color="blue"> at wt </font>: 
  
$$\tau_{4W}=E[Y_{wt,4W}]-E[Y_{wt,E16}]$$
And its **`lm` estimate**, $\hat{\tau}_{4W}$, is the *difference* of respective sample means (check below)

```{r,echo=FALSE,fig.height= 3.5, dev='svg'}
lucky +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
#circles around mean
  geom_point(aes(x = 1, y = mu.hat[1,2]),size=7,shape=1,colour='#00BFC4')+
  geom_point(aes(x = 2, y = mu.hat[3,2]),size=7,shape=1,colour='#00BFC4')+
#parameter and segments
  geom_text(aes(x = 2.2, y = abs(mu.hat[1,2]+mu.hat[3,2])/2, label = TeX("$\\widehat{\\tau_{4W}}$", output = "character")),colour="black", size=6, parse = TRUE) + 
  geom_segment(aes(x=2.1,y=mu.hat[1,2],xend=2.1,yend=mu.hat[3,2]),colour=1,arrow = arrow(length = unit(0.1,"cm"))) + #other way for double arrow??
  geom_segment(aes(x=1,y=mu.hat[1,2],xend=2,yend=mu.hat[1,2]),colour='grey',linetype=2)
```

```{r}
summary(twoFactFit)$coeff
means.2Fact
```

### Interaction effect: can we add up both effects??

$$Y_{ijk}=\theta + \tau_{KO} \times x_{KO,ijk}+ \tau_{4W} \times x_{4W,ijk}+ \tau_{KO4W} \times x_{KO,ijk}\times x_{4W,ijk}+ \varepsilon_{ijk}$$
  
Mathematically more challenging (try it!):
  
For any KO sample at 4w: $x_{KO}=1$ and $x_{4W}=1$. Then <font color="red"> $E[Y_{NrlKO,4W}]=\theta + \tau_{KO} + \tau_{4W} + \tau_{KO4W}$ </font>
  
Using the expectations from previous pages, you can show that:
  
$$\tau_{KO4W}=(E[Y_{NrlKO,E16}]-E[Y_{wt,E16}]) + (E[Y_{NrlKO,4W}]-E[Y_{wt,4W}])$$

```{r,echo=FALSE,fig.height= 3, dev='svg'}
mu.add<-mu.hat[3,2]-(mu.hat[1,2]-mu.hat[2,2])
lucky +
     scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
  #circles at means    
     geom_point(aes(x = 2, y = mu.hat[3,2]),size=7,shape=1,colour='#00BFC4')+
     geom_point(aes(x = 2, y = mu.hat[3,2]-(mu.hat[1,2]-mu.hat[2,2])),size=7,shape=1,colour='grey')+
  #parameters and segments
  #additive position
    geom_segment(aes(x=1,y=mu.hat[2,2],xend=2,yend=mu.add),colour='grey',linetype=2)+
  #tau_4W
  geom_segment(aes(x=2.1,y=mu.hat[3,2],xend=2.1,yend=mu.add),colour='grey',arrow = arrow(length = unit(0.1,"cm")))+
  geom_text(aes(x = 2.2, y = abs(mu.hat[3,2]+mu.add)/2, label = TeX("$\\widehat{\\tau_{4W}}$", output = "character")),colour='grey', size=6, parse = TRUE) +
  #interaction
  geom_segment(aes(x=2.1,y=mu.add,xend=2.1,yend=mu.hat[4,2]),colour='blue',arrow = arrow(length = unit(0.1,"cm"))) +
  geom_text(aes(x = 2.2, y = abs(mu.hat[4,2]+mu.add)/2, label = TeX("$\\widehat{\\tau_{KO4W}}$", output = "character")),colour="blue", size=6, parse = TRUE) 
```


```{r}
summary(twoFactFit)$coeff
means.2Fact
((means.2Fact$cellMeans[4]-means.2Fact$cellMeans[3])-
    (means.2Fact$cellMeans[2]-means.2Fact$cellMeans[1]))
```

## Two-way ANOVA without interaction: additive models

The interpretation of the coefficients changed when we drop the interaction terms

### Mathematically

$$Y_{ijk}=\theta + \tau_{KO} \times x_{KO,ijk}+ \tau_{4W} \times x_{4W,ijk} + \varepsilon_{ijk}$$
Note that this model is simpler and it has fewer parameters! but what do these mean?

As before, let's look at the expectations of each group:

- for any sample $i$ with $j=wt$ and $k=E16$: $x_{KO}=0$ and $x_{4W}=0$ (only $\theta$ remains):  $$E[Y_{wt,E16}]=\theta$$

- for any sample $i$ with $j=wt$ and $k=4W$: $x_{KO}=0$ and $x_{4W}=1$:  $$E[Y_{wt,4W}]=\theta + \tau_{4W}$$

- for any sample $i$ with $j=NrlKO$ and $k=E16$: $x_{KO}=1$ and $x_{4W}=0$:  $$E[Y_{KO,E16}]=\theta + \tau_{KO}$$

- for any sample $i$ with $j=NrlKO$ and $k=4W$: $x_{KO}=1$ and $x_{4W}=1$:  $$E[Y_{KO,4W}]=\theta + \tau_{KO} + \tau_{4W}$$

After some simple algebra, you get:

$$(E[Y_{wt,4W}]-E[Y_{wt,E16}]) + (E[Y_{KO,4W}]-E[Y_{KO,E16}]) = 2 \tau_{4W}$$

Then,

$$\tau_{4W}= (\text{Eff}_{4W|wt} + \text{Eff}_{4W|KO})/2$$
is the average effect of 4W over the levels of `gType`!!

Similar for the other *main effect*. 

## Some additional examples

### Example 4: development @wt is statistically significant

Again, the interaction effect is not significant, so there may be a developmetn effect *regardless* of the genotype. You need to test that hypothesis (main effects) using `anova` 


```{r,echo=FALSE}
egDat<-prDat %>% subset(row.names(prDat) %in%
       c("1447988_at","1438764_at")) %>%
       tibble::rownames_to_column(var = "gene")%>% 
       gather(sidChar, gExp,-gene) %>%
       inner_join(prDes,by="sidChar")%>% 
       mutate(grp=interaction(gType, devStage))
```

```{r,echo=FALSE}
multFit <- lm(gExp ~ gType * devStage, subset(egDat,gene=="1447988_at"))
summary(multFit)$coeff
```

```{r echo=FALSE, fig.height=4, fig.width=14,warning=FALSE,dev='svg'}
plot1Dat <- filter(egDat, gene == "1447988_at")
plot2Dat <- filter(egDat, gene == "1438764_at")

#cell-means for plot 1
mu.hat1<-plot1Dat %>% group_by(grp) %>% summarize(mean(gExp)) %>% as.data.frame()

plot1Lim <- ggplot(plot1Dat, aes(x = devStage, y = gExp, group=gType, colour=gType)) + 
  geom_jitter(width = 0.1) +
             labs(title = "1447988_at") +
             theme_bw() +
             theme(legend.position = "none") +
             ylim(5.5, 8) +  stat_summary(aes(group=gType,colour=gType), fun.y=mean, geom="line",size=1.5) +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
  #geom_segment(aes(x=1,y=mu.hat1[3,2],xend=1,yend=mu.hat1[1,2]),colour=1,arrow = arrow(length = unit(0.1,"cm")))+
  geom_segment(aes(x=1,y=mu.hat1[1,2],xend=1,yend=mu.hat1[3,2]),colour=1,arrow = arrow(length = unit(0.1,"cm")))+ #other way for double arrow??
  geom_segment(aes(x=1,y=mu.hat1[3,2],xend=2,yend=mu.hat1[3,2]),colour='grey',linetype=2)

#cell-means for plot 2
mu.hat2<-plot2Dat %>% group_by(grp) %>% summarize(mean(gExp)) %>% as.data.frame()

plot2Lim <- ggplot(plot2Dat, aes(x = devStage, y = gExp, group=gType, colour=gType)) + 
  geom_jitter(width = 0.1) +
             labs(title = "1438764_at") +
             theme_bw() +
             theme(legend.position = "none") +
             ylim(5.5, 8) +  stat_summary(aes(group=gType,colour=gType), fun.y=mean, geom="line",size=1.5) +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))+
#  geom_segment(aes(x=1,y=mu.hat2[3,2],xend=1,yend=mu.hat2[1,2]),colour=1,arrow = arrow(length = unit(0.1,"cm")))+
  geom_segment(aes(x=1,y=mu.hat2[1,2],xend=1,yend=mu.hat2[3,2]),colour=1,arrow = arrow(length = unit(0.1,"cm")))+ #other way for double arrow??
  geom_segment(aes(x=1,y=mu.hat2[3,2],xend=2,yend=mu.hat2[3,2]),colour='grey',linetype=2)
grid.arrange(plot1Lim, plot2Lim, ncol = 2)
```


### Example 5: both simple development and genotype are statistically significant

but not the interaction... note the almost parallel pattern

```{r,echo=FALSE}
egDat<-prDat %>% subset(row.names(prDat) %in%
       c("1456930_at","1417810_a_at")) %>%
       tibble::rownames_to_column(var = "gene")%>% 
       gather(sidChar, gExp,-gene) %>%
       inner_join(prDes,by="sidChar")
```


```{r,echo=FALSE,highlight.output = c(3:5),warning=FALSE}
multFit <- lm(gExp ~ gType * devStage, subset(egDat,gene=="1456930_at"))
summary(multFit)$coeff
```

```{r echo=FALSE, fig.height=4, fig.width=14,warning=FALSE, dev='svg'}
plot1Dat <- filter(egDat, gene == "1456930_at")
plot2Dat <- filter(egDat, gene == "1417810_a_at")

plot1Lim <- ggplot(plot1Dat, aes(x = devStage, y = gExp, group=gType, colour=gType)) + 
  geom_jitter(width = 0.1) +
             labs(title = "1456930_at") +
             theme_bw() +
             theme(legend.position = "none") +
             ylim(8, 13) +  stat_summary(aes(group=gType,colour=gType), fun.y=mean, geom="line",size=1.5) +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))

plot2Lim <- ggplot(plot2Dat, aes(x = devStage, y = gExp, group=gType, colour=gType)) + 
  geom_jitter(width = 0.1) +
             labs(title = "1417810_a_at") +
             theme_bw() +
             theme(legend.position = "none") +
             ylim(8,13) +  stat_summary(aes(group=gType,colour=gType), fun.y=mean, geom="line",size=1.5) +
  scale_color_manual(values = c('wt' = '#00BFC4', 'NrlKO' = '#F8766D'))

grid.arrange(plot1Lim, plot2Lim, ncol = 2)
```

---
